Measuring matter effects in Neutron star - Black hole binaries with Advanced LIGO

### 2016-04-06 Paper/paper.tex
author name just 2
updating intro

\newcommand{\Caltech}{\affiliation{Theoretical Astrophysics 350-17,
    California Institute of Technology, Pasadena, CA 91125, USA}}
\newcommand{\Cornell}{\affiliation{Center for Radiophysics and Space
    Research, Cornell University, Ithaca, New York 14853, USA}}
\newcommand{\CITA}{\affiliation{Canadian Institute for Theoretical
    Astrophysics, 60 St.~George Street, University of Toronto,
    Toronto, ON M5S 3H8, Canada}} %
\newcommand{\GWPAC}{\affiliation{Gravitational Wave Physics and
    Astronomy Center, California State University Fullerton,
    Fullerton, California 92834, USA}} %

@@

\author{People across the Atlantic}

@@

\begin{abstract}
Very recently, the Advanced LIGO (aLIGO) observatories announced the first
terrestrial detection of gravitational waves. This heralds an era of 
observational gravitational wave (GW) astronomy. Neutron star - black hole
(NSBH) binaries are one of the primary sources targeted for observation
with the aLIGO detectors.
% 
GWs from these sources will carry signatures of the tidal distortion of
the NS by the BH during the inspiral, and of its tidal disruption (if that
happens) at merger. In the NS is disrupted, the merger and ringdown signal
is drastically suppressed, which distinguishes it from a BHBH merger.
% 
In this paper, we present a Bayesian study of the measurability of the 
tidal deformability of neutron stars, allowing the companion hole to take 
on a variety of masses and spins.
% 
First, we find that if BHBH templates are used to estimate source 
parameters for an NSBH signal, a significant bias is introduced in the 
inference if the signal-to-noise ratio (SNR) is greater than $\rho=30$.
% 
Second, we find that if we are to get a loud signal with $\rho\geq 23$, 
we can put a factor of two bound on the NS tidal deformability parameter
$\lambdans\propto (R/M)^{5}$.
% 
Finally, we study the improvement in our measurement of $\lambdans$ from 
multiple observations. Using an approximately astrophysical population, 
we find that (i) the median measured $\lambdans$ catches up to within
$10\%$ of the true value in approximately $20$ observations; (ii) the $90\%$
confidence interval for $\lambda$ shrinks to $\pm 50\%$ within the same
number of observatoins.
% 
We find these results encouraging, and recommend that an effort to 
measure $\lambdans$ be taken within the LIGO-Virgo Collaboration once
we begin to detect NSBH coalescences.
\end{abstract}

@@

The Advanced LIGO (aLIGO) observatories began their first observing run ``O1''
mid-September 2015, operating at a factor of $3-4$ higher gravitational-wave (GW) 
strain sensitivity than their first-generation counterparts~\cite{Shoemaker2009}.
Coalescing binaries of
compact objects, i.e. Black Holes (BH) and/or Neutron Stars (NS), are the most
likely targets for GW detection with these ground-based instruments. Over a
$3$-month O1, these detectors would have access to as
many binary mergers as initial-LIGO detectors did over approximately $6$ years
of coincident operation. Both of these instruments will be further upgraded in
phases to reach their design sensitivity by $2018-19$, at which point we expect to
observe $\sim 70$ binary mergers a year~\cite{Abadie:2010cf}.

NS-BH binaries are one of the most important sources of gravitational waves (GW)
with these ground-based instruments, and we expect to detect $\mathcal{O}(10)$
of such systems every year~\cite{Abadie:2010cf}. 
%
One question that aLIGO will likely probe is `what is the nature of
matter at nuclear densities', i.e. what is the equation of state (EoS) of 
neutron stars. The EoS determines the compactness of NSs, which determines
their deformation over the course of inspiral (albeit weakly, at $5$PN order in 
phasing~\prayush{CITE PN TIDAL RESULTS}), as well as how early in the coalescence
process will the NS be disrupted by the tidal field of the BH~\prayush{[CITE]}.
Once disrupted,
the NS forms an accretion disk around the BH. Most of this matter disk spirals 
into the BH within a few milliseconds, with a small fraction becoming unbounded
and gets ejected~\prayush{CITE BHNS NR}. It is believed that an accompanying
short (lasting less than $2$ seconds) Gamma-ray Burst (sGRB) might be present
under certain physical conditions~\prayush{CITE NSBH-GRB},
making these systems one of the prime targets for joint electromagnetic-GW
astronomy searches~\prayush{CITE EM-GW}.
In such a situtation, the emitted GW signal
is heavily damped post-disruption, and the quasi-normal modes of the resulting
BH are not excited. Such a GW signal is qualitatively very different from a
BH-BH merger, and one of the main goals of this work is to determine the 
signal strength at which this difference will be quantifiabe by aLIGO.
%
However, studies have shown that for mass-ratio $\mbh/\mns\gtrsim 5$ the BH
will need to have fairly large and positively-aligned (with the orbit) spins
in order for its tidal field to be strong enough to disrupt the NS before the
innermost stable circular orbit (ISCO) is reached~\cite{Foucart:2013psa}.

>>>>

\newcommand{\Caltech}{\affiliation{Theoretical Astrophysics 350-17,
    California Institute of Technology, Pasadena, CA 91125, USA}}
\newcommand{\Cornell}{\affiliation{Center for Radiophysics and Space
    Research, Cornell University, Ithaca, New York 14853, USA}}
\newcommand{\CITA}{\affiliation{Canadian Institute for Theoretical
    Astrophysics, 60 St.~George Street, University of Toronto,
    Toronto, ON M5S 3H8, Canada}} %
\newcommand{\GWPAC}{\affiliation{Gravitational Wave Physics and
    Astronomy Center, California State University Fullerton,
    Fullerton, California 92834, USA}} %
\newcommand{\AEI}{\affiliation{Albert Einstein Institute,
Am M\"uhlenberg, G\"olm, Germany}} %

@@

\author{Prayush Kumar}\CITA\email{prkumar@cita.utoronto.ca}
\author{Michael P\"urrer}\AEI
\author{People across the Atlantic}

@@

\begin{abstract}
Very recently, the Advanced LIGO (aLIGO) observatories announced the first
terrestrial detection of gravitational waves. This heralds an era of 
observational gravitational wave (GW) astronomy. Neutron star - black hole
(NSBH) binaries are one of the primary sources targeted for observation
with the aLIGO detectors.
% 
GWs from these sources will carry signatures of the tidal distortion of
the NS by the BH during the inspiral, and of its tidal disruption (if that
happens) at merger. In the NS is disrupted, the merger and ringdown signal
is drastically suppressed, which distinguishes it from a BHBH merger.
% 
In this paper, we present the first Bayesian study of the measurability of
the tidal deformability of neutron stars, allowing the companion hole to take 
on a variety of masses and spins.
% 
First, we find that if BHBH templates are used to estimate source 
parameters for an NSBH signal, a significant bias is introduced in the 
inference if the signal-to-noise ratio (SNR) is greater than $\rho\simeq30$.
% 
Second, we find that if we are to get a loud signal with $\rho\geq 23$, 
we can put a factor of two bound on the NS tidal deformability parameter
$\lambdans\propto (R/M)^{5}$.
% 
Finally, we study the improvement in our measurement of $\lambdans$ from 
multiple observations. Using an approximately astrophysical population, 
we find that (i) the median measured $\lambdans$ catches up to within
$10\%$ of the true value in approximately $20$ observations; (ii) the $90\%$
confidence interval for $\lambdans$ shrinks to $\pm 50\%$ around the true 
value with the same number of observations.
% 
We find these results encouraging, and recommend that an effort to 
measure $\lambdans$ be taken within the LIGO-Virgo Collaboration once
we begin to detect NSBH coalescences.
\end{abstract}

@@

The Advanced LIGO (aLIGO) observatories began their first observing run ``O1''
mid-September 2015, operating at a factor of $3-4$ higher gravitational-wave (GW) 
strain sensitivity than their first-generation counterparts~\cite{Shoemaker2009}.
% 
During O1, the first terrestrial detection of gravitational waves was made, as
was announced earlier in 2016~\cite{LIGOVirgo2016a}. Emitted more than a billion
year ago, these waves herald the era of observational gravitational wave (GW)
astronomy as they pass through Earth.
% 
Over a $3$-month O1, aLIGO had access to more binary mergers than initial-LIGO
over its entire coincident operation period. Both of aLIGO instruments will be
further upgraded in phases to reach their design sensitivity by $2018-19$, at 
which point we may expect to observe $\sim 70$ binary mergers a 
year~\cite{Abadie:2010cf}.

Coalescing binaries of compact objects, i.e. black holes (BH) and/or neutron 
stars (NS), remain the most interestings targets for GW based observation with
aLIGO. NSBH binaries form an important class of compact binaries, and we expect
to see $\mathcal{O}(10)$ of these every year with design-sensitivity 
aLIGO~\cite{Abadie:2010cf}.
% 
NSBH mergers are of interest for various reasons. For instance, they have 
been associated with (as the progenitors of) short Gamma-ray Bursts
(GRBs)~\cite{eichler:89,1992ApJ...395L..83N,moch:93,Barthelmy:2005bx,
2005Natur.437..845F,2005Natur.437..851G,Shibata:2005mz,Paschalidis2014,
Tanvir:2013}. Plausible mechanisms have been proposed for such a 
scenario~\cite{Paschalidis2014,Ciolfi:2014yla}. A joint detection of GWs
from an NSBH system with a GRB would validate this hypothesis.
% 
Another question that NSBH mergers can be used to probe is `what is the 
nature of matter at nuclear densities', i.e. what is the equation of state
(EoS) of neutron stars. 
% 
The gravitational-waves emitted by NSBH binaries carry subtle hints of the 
matter structure of the neutron star. During early inspiral, the (weak) tidal
field of the companion hole deforms the star, thereby exciting its resonant
oscillation modes. This drains energy from the orbit resulting in an
increase of the inspiral rate. We know the leading and next-to-leading
order terms in post-Newtonian (PN) theory that capture this 
effect~\cite{Vines2011}, entering binary phasing at $5$PN. Closer to merger, 
the neutron star could get disrupted by the tidal field of its companion.
Once disrupted, the neutron star material may form an accretion disk around 
the black hole. Most of this matter disk spirals in within a few 
milliseconds, with a small fraction becoming unbounded and gets 
ejected~\cite{Foucart2015,Lovelace:2013vma,Deaton2013,Foucart2012,
Paschalidis2014} (under favorable conditions, and possible accompanied by 
gamma-ray emission).
The emitted GW signal is, as a result, heavily damped post-disruption, 
and the quasi-normal modes of the resulting black hole are not excited. Such a
GW signal will be qualitatively very different from that from an equivalent
binary black hole (BBH) merger. The primary goal of this work is to understand
what it requires to optimally extract information about neutron star structure
from such NSBH signals.

However, studies have shown that for mass-ratio $\mbh/\mns\gtrsim 5$ the BH
will need to have fairly large and positively-aligned (with the orbit) spins
in order for its tidal field to be strong enough to disrupt the NS before the
innermost stable circular orbit (ISCO) is reached~\cite{Foucart:2013psa}.
% 
It is worth noting
that not all point-particle terms in PN theory are known up to such high order,
and therefore we need to be careful when using PN waveforms to measure tidal
effects.
%
